---
title: "RTbySpeedAnalysis"
output: html_document
---

This documents shows how Reaction Times (RT) to a potential hazard may vary depending on driving speed. 
The dataset presents two speed categories: highway (1) and city (0). 
A potential hazard is represented by a pedestrians standing on the sidewalk at 4 or 14 degrees eccentricity.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

require(ggplot2)
require(GGally) #ggpirs
require(reshape2) #melt
#for mixed effects models
require(lme4)
require(optimx)
require(LMERConvenienceFunctions)
require(boot)
require(lattice)

#for scatterplot
require(car)

#these 3 are for correlations
require(Hmisc)
require(ggm)
require(psych)
require(gdata)
require(rJava)
require(Rmisc)

library(grid)
library(plyr)

require(nlme) #lme This generic function fits a linear mixed-effects model in the formulation described in Laird and Ware (1982) but allowing for nested random effects. The within-group errors are allowed to be correlated and/or have unequal variances.
require(multcomp)
require(km.ci)


#Set path and read file          
#setwd("AnalysisDemo")
originalmydataold <-read.table('MinimumDistanceNearPedsScores_stationary_biopeds4.txt', header = TRUE, sep = "")
originalmydata<- data.frame(originalmydataold)
lapply(originalmydata, class)

#Explain Coding in Variables
#conditionstatic is 0 for walking pedestrians and 1 for stationary pedestrians


#only include straight driving segments
wordList1 <- c("turn","curve", "grabber")
wordList2 <- c("turn","curve")
mydata1 <-subset(originalmydata, conditionstatic == 1 & !( popUpType %in% wordList1)) #for stationar we exclude list1
mydata2 <-subset(originalmydata, conditionstatic == 0 & !( popUpType %in% wordList2)) #for biopeds we exclude list2
mydata <- rbind (mydata1, mydata2)

#Recode variables
mydata$timely[mydata$timely=="-1"] <- 3 #misses
mydata$timely[mydata$timely=="-2"] <- 3 #misses
mydata$timely[mydata$timely=="0"] <- 2 #late
mydata$timely[mydata$timely=="1"] <- 1 #timely

mydata$pedEccentricity[mydata$pedEccentricity=="-4"] <- 4
mydata$pedEccentricity[mydata$pedEccentricity=="-14"] <- 14
mydata$pedEccentricity[mydata$pedEccentricity=="4"] <- 4
mydata$pedEccentricity[mydata$pedEccentricity=="14"] <- 14

#Calculate new variables and concatenate
RT = mydata$honkTime-mydata$warpTime
mydatamodule <- cbind (mydata, RT)


#Subset datasets independently for the seeing side and the blindside
mydataBLINDLARGE <-subset(mydatamodule, pedside == 1 & pedEccentricity == 14)
dim(mydataBLINDLARGE)
mydataSELARGE <-subset(mydatamodule, pedside == 0 & pedEccentricity == 14)
dim(mydataSELARGE)
mydataBLINDsmall <-subset(mydatamodule, pedside == 1 & pedEccentricity == 4)
dim(mydataBLINDsmall)
mydataSEsmall <-subset(mydatamodule, pedside == 0 & pedEccentricity == 4)
dim(mydataSEsmall)


# function for number of observations 
give.n <- function(x){
  return(c(y = median(x)*1.05, label = length(x))) 
  # experiment with the multiplier to find the perfect position
}

#analyze the small and large eccentricity pedestrians on the seeing side
dataset<- rbind(mydataSELARGE, mydataSEsmall)

#only include rows with possible Speed values
datasetpolished<- subset(dataset, dataset$SpeedAtMin >0 & !is.na(dataset$SpeedAtMin))

#only include events when the pedestrian was detected
DatasetTimelyLate<- subset(datasetpolished, datasetpolished$timely ==1 |datasetpolished$timely ==2)

#assign variables names for plots
PedEcc <-as.factor(DatasetTimelyLate$pedEccentricity)
SpeedCat<- as.factor(DatasetTimelyLate$speedcat)
ReactionTime<-DatasetTimelyLate$RT

#Build Boxplot of Pedestrian Eccentricity for Speed Category
subset1 <- data.frame(DatasetTimelyLate$pedEccentricity, DatasetTimelyLate$speedcat, DatasetTimelyLate$RT)
	tmp <- melt(subset1, id=c( "DatasetTimelyLate.pedEccentricity", "DatasetTimelyLate.speedcat"))
	SpeedCatForpedEccentricity<-ggplot(tmp, aes(factor( DatasetTimelyLate$speedcat), y = value,  fill=as.factor( DatasetTimelyLate$pedEccentricity))) +
  geom_boxplot() + 
  facet_wrap(~variable) +
  stat_summary(fun.data = give.n, geom = "text", fun.y = median, vjust = 1) +
  theme(legend.title = element_blank())+
	  scale_x_discrete( breaks=c("0", "1"), labels=c("city", "highway"))+
	  scale_colour_manual(breaks=c("4", "14"), labels=c("4", "14")) +
	  labs(y="Reaction Times (secs)", x = "Speed Category")



#create contrast matrix
modellmint<-lm(RT ~ as.factor(pedEccentricity) *as.factor(speedcat), data=DatasetTimelyLate)
#
group <- paste0(as.factor(DatasetTimelyLate$pedEccentricity), as.factor(DatasetTimelyLate$speedcat))
#pastes after converting into character class
modelmatrix<-model.matrix(modellmint)
#creates a design matrix by expanding factors to a set dummy variables
group <- aggregate(modelmatrix ~ group, FUN=mean)
#splits data into subsets, computes summary statistics (used to create a short model matrix)
rownames(group) <- group$group #changes rownames from incremental numbers to group variable
(group <- group[,-1]) #delete the column group, which is now rownames




#Created point plots with Confidence Intervals of the mean: Pedestrian Eccentricity by SpeedCat 
subset2 <- data.frame(PedEcc, SpeedCat, ReactionTime)

tgc <- summarySE(subset2, measurevar="ReactionTime", groupvars=c("PedEcc", "SpeedCat"))

ReactionTimesPedEccbySpeedCat1<- ggplot(tgc, aes(PedEcc, y= ReactionTime, pch=SpeedCat)) +
	geom_errorbar(aes(ymin=ReactionTime-ci, ymax=ReactionTime+ci, colour=SpeedCat), size=  0.8,width=.1, position=position_dodge(.2)) +
		geom_point(aes(colour=SpeedCat, shape=SpeedCat),size=  2, stat = "identity", position=position_dodge(.2)) +  ylim(0, 5) + 
	ggtitle("Reaction Time")+ labs( y= "Reaction Time (sec)") +
	scale_shape_manual(values=c(19,20),  breaks=c("0", "1"), labels=c("city", "highway")) +
      scale_colour_manual(values=c("black", "red"),  breaks=c("0", "1"), labels=c("city", "highway")) +
	scale_linetype_manual(values=c(1,2),  breaks=c("0", "1"), labels=c("city", "highway")) 

ReactionTimesPedEccbySpeedCat<-ReactionTimesPedEccbySpeedCat1+theme(panel.background = element_rect(fill = 'white', colour = 'grey20'), text = element_text(size = 22))+ scale_fill_manual(values=c("#999999", "black"), 
                         breaks=c("0", "1"), labels=c("city", "highway"))+theme_classic()+ scale_x_discrete( breaks=c("4", "14"), labels=c("4", "14"))

#Created point plots with Confidence Intervals of the mean: SpeedCat by Pedestrian Eccentricity
ReactionTimesSpeedCatbyPedEcc1<- ggplot(tgc, aes(SpeedCat, y= ReactionTime, pch=PedEcc)) +
	geom_errorbar(aes(ymin=ReactionTime-ci, ymax=ReactionTime+ci, colour=PedEcc), size=  0.8,width=.1, position=position_dodge(.2)) +
	geom_line(data=tgc, aes(x=SpeedCat, y=ReactionTime, colour=PedEcc, linetype = PedEcc), size=  0.8,position=position_dodge(.2)) +
	geom_point(aes(colour=PedEcc, shape=PedEcc),size=  2, stat = "identity", position=position_dodge(.2)) +  ylim(0, 5) + 
 	labs( y= "Reaction Times") +
	scale_shape_manual(values=c(19,19,19),  breaks=c("4", "14"), labels=c("4", "14")) +
      scale_colour_manual(values=c("#999999", "cyan", "blue"),  breaks=c("4", "14"), labels=c("4", "14")) +
	scale_linetype_manual(values=c(1,1,1),  breaks=c("4", "14"), labels=c("4", "14")) 

ReactionTimesSpeedCatbyPedEcc<-ReactionTimesSpeedCatbyPedEcc1+theme(panel.background = element_rect(fill = 'white', colour = 'grey20'), text = element_text(size = 22))+ scale_fill_manual(values=c("cyan", "blue"), breaks=c("4", "14"), labels=c("4", "14"))+theme_classic() +
				 labs( y= "Reaction Times (secs)") +
				scale_x_discrete( breaks=c("0", "1"), labels=c("City", "Highway"))






#adding Clustering analysis to split data in 2 groups, with scaled data
dk<-cbind(DatasetTimelyLate$RT, DatasetTimelyLate$SpeedAtMin)
dkscaled<-scale(dk)
classRT<- kmeans(dkscaled, 2, nstart=25)  
DatasetTimelyLate<-cbind(DatasetTimelyLate,classRT$cluster )


#Build Scatterplot showing the 2 Clusters 
subset <- data.frame(DatasetTimelyLate$speedcat, DatasetTimelyLate$timely, DatasetTimelyLate$RT, classRT$cluster)
tmp <- melt(subset, id=c( "DatasetTimelyLate.speedcat", "DatasetTimelyLate.timely", "classRT.cluster"))

ScatterCluster<-ggplot(DatasetTimelyLate, aes(x=SpeedAtMin, y=RT, color=as.factor(classRT$cluster))) + 
  geom_point(shape=1)+
  labs(y="Reaction Times (secs)", x = "Speed")+
  scale_color_manual(name = "Cluster",labels = c("1", "2"), values = c("black", "magenta"))

#Build Scatterplot showing the Speed Category Split
ScatterSpeedCategory<-ggplot(DatasetTimelyLate, aes(x=SpeedAtMin, y=RT, color=as.factor(speedcat))) + 
  geom_point(shape=1)+
  labs(y="Reaction Times (secs)", x = "Speed")+
  scale_color_manual(name = "Speed Category",labels = c("City", "Highway"), values = c("purple", "green"))

#Build Scatterplot showing the Potential Collision Split
ScatterCollision<-ggplot(DatasetTimelyLate, aes(x=SpeedAtMin, y=RT, color=as.factor(timely))) + 
  geom_point(shape=1)+
  labs(y="Reaction Times (secs)", x = "Speed")+
  scale_color_manual(name = "Potential Collision",labels = c("NO", "YES"), values = c("blue", "red"))
  


#Build Boxplot showing the the Potential Collisions by Clusters distributions
CollisionbyCluster<-ggplot(tmp, aes(factor( DatasetTimelyLate.timely), y = value,  fill=as.factor( classRT.cluster ))) +
  geom_boxplot() + 
  facet_wrap(~variable) +
  stat_summary(fun.data = give.n, geom = "text", fun.y = median, vjust = 1) +
 # theme(legend.title = element_blank())+
   scale_x_discrete( breaks=c("1", "2"), labels=c("NO", "YES"))+
  scale_fill_discrete(name = "Cluster", breaks=c("1", "2"), labels=c("Low Speed", "High Speed"))+
                        labs(y="Reaction Times (secs)", x = "Potential Collision")



```


## Descriptives
Reaction Times 
```{r dataset)}
summary(DatasetTimelyLate$RT)
SpeedCatForpedEccentricity
##dataset$scenario
```




## Statistical Analysis
```{r dataset, echo=TRUE, message=FALSE, warning=FALSE}

#test <- lm  (RT ~ as.factor(speedcat), data = DatasetTimelyLate)
#summary(test)

test <- lm  (RT ~ as.factor(pedEccentricity) *as.factor(speedcat), data = DatasetTimelyLate)
summary(test)
anova(test)

test <- lme  (RT ~ as.factor(pedEccentricity) *as.factor(speedcat), random= ~1 | subject, data=DatasetTimelyLate)
summary(test)
anova(test)

rbind(group["40",] - group["140",], #4 City  vs  14 City
      group["41",] - group["141",]) #4 Highway vs 14 Highway 

contrast.matrix <- rbind(
  `4:city vs. 14:city` = c(0, -1, 0, 0),
  `4:highway vs. 14:highway`     = c(0, -1, 0, -1))

summary(glht(test, contrast.matrix), test = adjusted("bonferroni"))

ReactionTimesPedEccbySpeedCat

rbind(group["40",] - group["41",], #4 City vs 4 Highway
      group["140",] - group["141",]) #14 City vs 14 Highway


contrast.matrix <- rbind(
  `4:city vs. 4:highway` = c(0, 0, -1, 0),
  `14:city vs. 14:highway`     = c(0, 0, -1, -1))

summary(glht(test, contrast.matrix), test = adjusted("bonferroni"))

ReactionTimesSpeedCatbyPedEcc

dk<-cbind(DatasetTimelyLate$RT, DatasetTimelyLate$SpeedAtMin)
dkscaled<-scale(dk)
classRT<- kmeans(dkscaled, 2, nstart=25)  

ScatterCluster

ScatterSpeedCategory

ScatterCollision


CollisionbyCluster





```


##Conclusions 1
RTs are larger in the High Speed Category (Highway) than in the Low Speed Category (City). 
RTs also depend on the eccentricity of the potential hazard: objects that are more peripheral are reported later than more central objects. 

Cluster Analysis successfully identified all the high speeds events.
Potential collisions fell within the High Speed cluster. 

